---
format:
  revealjs:
    output-file: 02-stan-ecosystem-slides.html
---

# The Stan Ecosystem {#sec-stan-ecosystem}

By the end of this section, you'll understand that the Stan ecosystem is a 
stack of related projects mainly developed under
[`stan-dev`](https://github.com/stan-dev), ranging from the compiler and math
library to `CmdStan`, language interfaces, and workflow packages. You'll be 
able to place a typical modeling task within that stack and identify which 
layer each piece belongs to.

---

## A minimal Stan workflow

We start our workflow by writing a minimal Stan model in a `.stan` file.

```{r}
#| eval: false
#| filename: "R · Stan model"

stan_code <- "
data {
  int<lower=0> N;
  vector[N] y;
}
parameters {
  real mu;
  real<lower=0> sigma;
}
model {
  mu ~ normal(0, 10);      // prior
  sigma ~ exponential(1);  // prior
  target += normal_lpdf(y | mu, sigma);  // likelihood
}
generated quantities {
  vector[N] log_lik;
  for (n in 1:N) {
    log_lik[n] = normal_lpdf(y[n] | mu, sigma);
  }
}
"
writeLines(stan_code, "model.stan")
```

---

Additionally, we provide some data for the input and save it as a JSON file.

```{r}
#| eval: false
#| filename: "R · data"
library(cmdstanr)

stan_data <- list(N = 20, y = rnorm(20, mean = 5, sd = 2))
cmdstanr::write_stan_json(stan_data, "data.json")

> stan_data
$N
[1] 20

$y
 [1] 4.543589 4.899750 4.081620 8.698614 4.827003 2.903813 8.659615 5.165155
 [9] 5.148127 1.206675 2.368544 7.873981 5.071258 5.322390 6.913431 7.336083
[17] 3.462279 3.717310 1.620141 3.577615
```

---

### Language interfaces

Next we can compile and sample from the Stan model. To do this, we use an
interface, such as [`cmdstanr`](https://mc-stan.org/cmdstanr/) for R or 
[`cmdstanpy`](https://mc-stan.org/cmdstanpy/) for Python, that wraps
[`CmdStan`](https://mc-stan.org/docs/cmdstan-guide/). 

<br>

`CmdStan` invokes

+ [`stanc3`](https://github.com/stan-dev/stanc3) to transpile the Stan model into a C++ file, which is then compiled and linked against 
+ the [`math`](https://mc-stan.org/math/) library (implementing built-in functions for distributions, linear algebra, ODE solvers, etc., along with automatic differentiation) and 
+ the [`stan`](https://github.com/stan-dev/stan) library (implementing the inference algorithms, such as HMC/NUTS). 

The result is a standalone executable that the interface then runs to produce
posterior samples.

---

#### The R interface [`cmdstanr`](https://mc-stan.org/cmdstanr/)

Compile and sample from the Stan model in R with `cmdstanr`.
Use the fitted model to extract posterior draws, plot MCMC trace plots with 
`bayesplot`, and run approximate leave-one-out cross-validation with `loo`.

```{r}
#| eval: false
#| filename: "R · cmdstanr"

library(cmdstanr)
library(bayesplot)
library(loo)

# compile Stan model
mod <- cmdstanr::cmdstan_model("model.stan")

# sample from the model
fit <- mod$sample(data = "data.json", chains = 4, iter_sampling = 1000)

# get draws
draws <- fit$draws(variables = c("mu", "sigma"))

# plotting and cross-validation
bayesplot::mcmc_trace(draws)
loo::loo(fit$draws("log_lik"))
```

---

#### The Python interface [`cmdstanpy`](https://mc-stan.org/cmdstanpy/)

Compile and sample from the Stan model in Python with `cmdstanpy`. Use the fitted model to extract posterior draws, plot MCMC trace 
plots, and run approximate leave-one-out cross-validation using `arviz`.

```{python}
#| eval: false
#| filename: "Python  · cmdstanpy"

from cmdstanpy import CmdStanModel
import arviz as az

# compile Stan model
mod = CmdStanModel(stan_file="model.stan")

# sample from the model
fit = mod.sample(data="data.json", chains=4, iter_sampling=1000)

# get draws
idata = az.from_cmdstanpy(fit, log_likelihood="log_lik")

# plotting and cross-validation
az.plot_trace(idata, var_names=["mu", "sigma"])
az.loo(idata)
```

---

#### The Command-line interface [`CmdStan`](https://mc-stan.org/docs/cmdstan-guide/)

`cmdstanr` and `cmdstanpy` are thin language wrappers around `CmdStan`, the
command-line interface to Stan. `CmdStan` itself invokes `stanc3` to compile
a `.stan` file into a standalone executable, which can be run directly from
the command line to sample from (or optimize, etc.) the model. 

```{bash}
#| eval: false
#| filename: "Bash"

# run from the CmdStan directory
make model

./model sample num_chains=4 num_samples=1000 \
  data file=data.json \
  output file=output.csv

$CMDSTAN/bin/stansummary output_*.csv
$CMDSTAN/bin/diagnose output_*.csv
```

---

### Higher-level modelling frameworks

Instead of writing a Stan model from scratch, you can use higher-level
modelling frameworks that let you specify your model using formula syntax.
Two example frameworks are 

+ [`brms`](https://paulbuerkner.com/brms/), which
dynamically generates and compiles custom Stan code for your specified model,
and 
+ [`rstanarm`](https://mc-stan.org/rstanarm/), which fits your model using
a set of pre-compiled Stan programs for common model families (avoiding
compilation time, at the cost of some flexibility).

:::: {.columns}

::: {.column width="50%"}

```{r}
#| eval: false
#| filename: "R · brms"

library(brms)

dat <- data.frame(y = jsonlite::fromJSON("data.json")$y)

fit <- brms::brm(
  formula = y ~ 1,
  data = dat,
  family = gaussian(),
  prior = c(
    prior(normal(0, 10), class = Intercept),
    prior(exponential(1), class = sigma)
  ),
  chains = 4,
  iter = 2000
)
```

:::

::: {.column width="50%"}

```{r}
#| eval: false
#| filename: "R · rstanarm"

library(rstanarm)

dat <- data.frame(y = jsonlite::fromJSON("data.json")$y)

fit <- rstanarm::stan_glm(
  formula = y ~ 1,
  data = dat,
  family = gaussian(),
  prior_intercept = normal(0, 10, autoscale = FALSE),
  prior_aux = exponential(1, autoscale = FALSE),
  chains = 4,
  iter = 2000
)
​
​
```

:::

::::

---

## The Stan ecosystem and its libraries
To summarize what we have seen so far, we can outline the different layers
in the Stan ecosystem and how they relate to one another.

::: {.callout-note}
## Different layers in the Stan ecosystem

+ **Layer 1: Compiler & math**: `stanc3` · `stan` · `math`
  + Contribute to the fundamental building blocks of the Stan language: compiler transpilation, autodiff, or core inference algorithms.
+ **Layer 2: Command-line engine**: `cmdstan`
  + Contribute to the command-line interface that drives compilation and execution.
+ **Layer 3: Language interfaces**: `cmdstanr` · `cmdstanpy` · `rstan` · ...
  + Contribute to environment-specific host language wrappers (R, Python, Julia, etc.) that interface with Stan.
+ **Layer 4: Workflow & analysis**: `posterior` · `bayesplot` · `loo` · `priorsense` · ...
  + Contribute to downstream packages used among others for visualization, posterior diagnostics, cross-validation, and sensitivity analysis.

:::

---

### Compiler, math, and CmdStan

| Package | Description | Relations | Language |
|---------|-------------|-----------|----------|
| [`math`](https://mc-stan.org/math/) | Autodiff library for distributions, linear algebra, and gradients | Used by generated model code | C++ |
| [`stanc3`](https://github.com/stan-dev/stanc3) | Compiler that translates `.stan` to C++ | Feeds `cmdstan` and `rstan` | OCaml |
| [`stan`](https://github.com/stan-dev/stan) | Inference algorithms and services (HMC/NUTS, etc.) | Built on `math`; used via `cmdstan` / `rstan` | C++ |
| [`cmdstan`](https://github.com/stan-dev/cmdstan) | Command-line engine to compile and run Stan models | Wraps `stanc3` + `stan` + `math`; backend for `cmdstanr` / `cmdstanpy` | C++ / CLI |

---

### Language interfaces

| Package | Description | Relations | Language |
|---------|-------------|-----------|----------|
| [`cmdstanr`](https://github.com/stan-dev/cmdstanr) | R wrapper around CmdStan; returns `CmdStanMCMC` | Calls `cmdstan`; used with `posterior`, `bayesplot`, `loo` | R |
| [`cmdstanpy`](https://github.com/stan-dev/cmdstanpy) | Python wrapper around CmdStan; returns `CmdStanMCMC` | Calls `cmdstan`; used with `arviz` | Python |
| [`rstan`](https://github.com/stan-dev/rstan) | R interface that embeds Stan in-process; returns `stanfit` | Alternative to CmdStan; used by `rstanarm` and `rstantools` packages | R |

---

### Workflow, modeling, and packaging

| Package | Description | Relations | Language |
|---------|-------------|-----------|----------|
| [`brms`](https://github.com/paul-buerkner/brms) | Flexible multilevel / distributional regression via formulas; returns `brmsfit` | Generates Stan; backends `cmdstanr` / `rstan`; used with `loo`, `bayesplot` | R |
| [`rstanarm`](https://github.com/stan-dev/rstanarm/) | Precompiled applied regression models; returns `stanreg` | Built on `rstan`; used with `loo`, `bayesplot`, `projpred` | R |
| [`posterior`](https://github.com/stan-dev/posterior/) | Tools to manipulate and summarize draws (R-hat, ESS, etc.) | Shared draws API for `cmdstanr` and analysis packages | R |
| [`bayesplot`](https://mc-stan.org/bayesplot/) | ggplot2 plots for MCMC diagnostics, PPC, and posteriors | Uses draws from interfaces / `posterior` | R |

---

### Workflow, modeling, and packaging (continued I)

| Package | Description | Relations | Language |
|---------|-------------|-----------|----------|
| [`loo`](https://github.com/stan-dev/loo/) | Approximate LOO-CV (PSIS-LOO), WAIC, and model weights | Needs pointwise `log_lik` from fitted models | R |
| [`arviz`](https://github.com/arviz-devs/arviz) | Diagnostics, plots, and model comparison in Python; returns `xr.DataTree` | Python counterpart to `posterior` / `bayesplot` / parts of `loo` | Python |
| [`priorsense`](https://github.com/n-kall/priorsense) | Prior and likelihood sensitivity analysis (power-scaling) | Uses `posterior`; related to PSIS methods in `loo` | R |
| [`projpred`](https://mc-stan.org/projpred/) | Projection predictive variable selection | Uses `stanreg` (`rstanarm`) or `brmsfit` (`brms`) as reference | R |

---

### Workflow, modeling, and packaging (continued II)

| Package | Description | Relations | Language |
|---------|-------------|-----------|----------|
| [`shinystan`](https://github.com/stan-dev/shinystan/) | Interactive Shiny GUI for MCMC diagnostics | Uses `stanfit` / `stanreg`; overlaps with `bayesplot` | R |
| [`rstantools`](https://github.com/stan-dev/rstantools/) | Tools to ship Stan models inside R packages; models typically return `stanfit` | Scaffolding used by packages such as `rstanarm` | R |
| [`posteriordb`](https://github.com/stan-dev/posteriordb) | Database of models, data, and reference posteriors | Shared resource for testing, teaching, and CI | R / Python / data |
