---
format:
  revealjs:
    output-file: 02-stan-ecosystem-slides.html
---

# The Stan Ecosystem {#sec-stan-ecosystem}

+ understanding Stan as an ecosystem
+ `stan-dev` is a GitHub organization that contains many repositories that work together at different levels

---

## A minimal Stan workflow

```{r}
#| eval: false
#| filename: "R · Stan model"

stan_code <- "
data {
  int<lower=0> N;
  vector[N] y;
}
parameters {
  real mu;
  real<lower=0> sigma;
}
model {
  mu ~ normal(0, 10);      // prior
  sigma ~ exponential(1);  // prior
  target += normal_lpdf(y | mu, sigma);  // likelihood
}
generated quantities {
  vector[N] log_lik;
  for (n in 1:N) {
    log_lik[n] = normal_lpdf(y[n] | mu, sigma);
  }
}
"
writeLines(stan_code, "model.stan")
```

---

Provide some data for the input

```{r}
#| eval: false
#| filename: "R · data"
library(cmdstanr)

stan_data <- list(N = 20, y = rnorm(20, mean = 5, sd = 2))
cmdstanr::write_stan_json(stan_data, "data.json")

> stan_data
$N
[1] 20

$y
 [1] 4.543589 4.899750 4.081620 8.698614 4.827003 2.903813 8.659615 5.165155
 [9] 5.148127 1.206675 2.368544 7.873981 5.071258 5.322390 6.913431 7.336083
[17] 3.462279 3.717310 1.620141 3.577615
```

---

### Language interfaces

#### Using R with [`cmdstanr`](https://mc-stan.org/cmdstanr/)

```{r}
#| eval: false
#| filename: "R · cmdstanr"

library(cmdstanr)
library(bayesplot)
library(loo)

# compile Stan model
mod <- cmdstanr::cmdstan_model("model.stan")

# sample from the model
fit <- mod$sample(data = "data.json", chains = 4, iter_sampling = 1000)

# get draws
draws <- fit$draws(variables = c("mu", "sigma"))

# plotting and cross-validation
bayesplot::mcmc_trace(draws)
loo::loo(fit$draws("log_lik"))
```

---

#### Using Python with [`cmdstanpy`](https://mc-stan.org/cmdstanpy/)

```{python}
#| eval: false
#| filename: "Python  · cmdstanpy"

from cmdstanpy import CmdStanModel
import arviz as az

# compile Stan model
mod = CmdStanModel(stan_file="model.stan")

# sample from the model
fit = mod.sample(data="data.json", chains=4, iter_sampling=1000)

# get draws
idata = az.from_cmdstanpy(fit, log_likelihood="log_lik")

# plotting and cross-validation
az.plot_trace(idata, var_names=["mu", "sigma"])
az.loo(idata)
```

---

#### The Command-line interface [`CmdStan`](https://mc-stan.org/docs/cmdstan-guide/)

`cmdstanr` and `cmdstanpy` are thin language wrappers around the `CmdStan` command-line interface.

You can use `CmdStan` directly in the command line:

```{bash}
#| eval: false
#| filename: "Bash"

# run from the CmdStan directory
make model

./model sample num_chains=4 num_samples=1000 \
  data file=data.json \
  output file=output.csv

$CMDSTAN/bin/stansummary output_*.csv
$CMDSTAN/bin/diagnose output_*.csv
```

Further processing of the output is usually done in Python or R.

---

### Higher-level modelling frameworks

:::: {.columns}

::: {.column width="50%"}
#### [`brms`](https://paulbuerkner.com/brms/)

Fits customized Bayesian multilevel and non-linear models in Stan.

```{r}
#| eval: false
#| filename: "R · brms"

library(brms)

dat <- data.frame(y = jsonlite::fromJSON("data.json")$y)

fit <- brms::brm(
  formula = y ~ 1,
  data = dat,
  family = gaussian(),
  prior = c(
    prior(normal(0, 10), class = Intercept),
    prior(exponential(1), class = sigma)
  ),
  chains = 4,
  iter = 2000
)

bayesplot::mcmc_trace(fit, pars = c("b_Intercept", "sigma"))
loo::loo(fit)
```
:::

::: {.column width="50%"}
#### [`rstanarm`](https://mc-stan.org/rstanarm/)

Fits standard Bayesian regression models using pre-compiled Stan code.

```{r}
#| eval: false
#| filename: "R · rstanarm"

library(rstanarm)

dat <- data.frame(y = jsonlite::fromJSON("data.json")$y)

fit <- rstanarm::stan_glm(
  formula = y ~ 1,
  data = dat,
  family = gaussian(),
  prior_intercept = normal(0, 10, autoscale = FALSE),
  prior_aux = exponential(1, autoscale = FALSE),
  chains = 4,
  iter = 2000
)



bayesplot::mcmc_trace(fit, pars = c("(Intercept)", "sigma"))
loo::loo(fit)
```
:::

::::

---

## The Stan ecosystem and its libraries

::: {.callout-note}
## Different layers in the Stan ecosystem

+ **Layer 1: Compiler & math**: `stanc3` · `stan` · `math`
  + Contribute to the fundamental building blocks of the language: compiler transpilation, C++ reverse-mode autodiff, or core inference algorithms.
+ **Layer 2: Command-line engine**: `cmdstan`
  + Contribute to the lightweight C++ command-line interface that drives compilation and execution.
+ **Layer 3: Language interfaces**: `cmdstanr` · `cmdstanpy` · `rstan`
  + Contribute to environment-specific host language wrappers (R, Python, Julia, shell) that pass data to CmdStan and manage process calls.
+ **Layer 4: Workflow & analysis**: `posterior` · `bayesplot` · `loo` · `priorsense` · `rstanarm` · ...
  + Contribute to downstream packages used among others for visualization, posterior diagnostics, cross-validation, and sensitivity analysis.

:::

---

### Compiler, math, and CmdStan

| Package | Description | Relations | Language |
|---------|-------------|-----------|----------|
| [`math`](https://mc-stan.org/math/) | Autodiff library for distributions, linear algebra, and gradients | Used by generated model code | C++ |
| [`stanc3`](https://github.com/stan-dev/stanc3) | Compiler that translates `.stan` to C++ | Feeds `cmdstan` and `rstan` | OCaml |
| [`stan`](https://github.com/stan-dev/stan) | Inference algorithms and services (HMC/NUTS, etc.) | Built on `math`; used via `cmdstan` / `rstan` | C++ |
| [`cmdstan`](https://github.com/stan-dev/cmdstan) | Command-line engine to compile and run Stan models | Wraps `stanc3` + `stan` + `math`; backend for `cmdstanr` / `cmdstanpy` | C++ / CLI |

---

### Language interfaces

| Package | Description | Relations | Language |
|---------|-------------|-----------|----------|
| [`cmdstanr`](https://github.com/stan-dev/cmdstanr) | R wrapper around CmdStan; returns `CmdStanMCMC` | Calls `cmdstan`; used with `posterior`, `bayesplot`, `loo` | R |
| [`cmdstanpy`](https://github.com/stan-dev/cmdstanpy) | Python wrapper around CmdStan; returns `CmdStanMCMC` | Calls `cmdstan`; used with `arviz` | Python |
| [`rstan`](https://github.com/stan-dev/rstan) | R interface that embeds Stan in-process; returns `stanfit` | Alternative to CmdStan; used by `rstanarm` and `rstantools` packages | R |

---

### Workflow, modeling, and packaging

| Package | Description | Relations | Language |
|---------|-------------|-----------|----------|
| [`brms`](https://github.com/paul-buerkner/brms) | Flexible multilevel / distributional regression via formulas; returns `brmsfit` | Generates Stan; backends `cmdstanr` / `rstan`; used with `loo`, `bayesplot` | R |
| [`rstanarm`](https://github.com/stan-dev/rstanarm/) | Precompiled applied regression models; returns `stanreg` | Built on `rstan`; used with `loo`, `bayesplot`, `projpred` | R |
| [`posterior`](https://github.com/stan-dev/posterior/) | Tools to manipulate and summarize draws (R-hat, ESS, etc.) | Shared draws API for `cmdstanr` and analysis packages | R |
| [`bayesplot`](https://mc-stan.org/bayesplot/) | ggplot2 plots for MCMC diagnostics, PPC, and posteriors | Uses draws from interfaces / `posterior` | R |

---

### Workflow, modeling, and packaging (continued I)

| Package | Description | Relations | Language |
|---------|-------------|-----------|----------|
| [`loo`](https://github.com/stan-dev/loo/) | Approximate LOO-CV (PSIS-LOO), WAIC, and model weights | Needs pointwise `log_lik` from fitted models | R |
| [`arviz`](https://github.com/arviz-devs/arviz) | Diagnostics, plots, and model comparison in Python; returns `DataTree` | Python counterpart to `posterior` / `bayesplot` / parts of `loo` | Python |
| [`priorsense`](https://github.com/n-kall/priorsense) | Prior and likelihood sensitivity analysis (power-scaling) | Uses `posterior`; related to PSIS methods in `loo` | R |
| [`projpred`](https://mc-stan.org/projpred/) | Projection predictive variable selection | Uses `stanreg` (`rstanarm`) or `brmsfit` (`brms`) as reference | R |

---

### Workflow, modeling, and packaging (continued II)

| Package | Description | Relations | Language |
|---------|-------------|-----------|----------|
| [`shinystan`](https://github.com/stan-dev/shinystan/) | Interactive Shiny GUI for MCMC diagnostics | Uses `stanfit` / `stanreg`; overlaps with `bayesplot` | R |
| [`rstantools`](https://github.com/stan-dev/rstantools/) | Tools to ship Stan models inside R packages; models typically return `stanfit` | Scaffolding used by packages such as `rstanarm` | R |
| [`posteriordb`](https://github.com/stan-dev/posteriordb) | Database of models, data, and reference posteriors | Shared resource for testing, teaching, and CI | R / Python / data |
