---
format:
  revealjs:
    output-file: 07-walkthrough-R-slides.html
---
# Walkthrough: Contributing to Stan's R packages

In this tutorial, we will walk you through one example code contribution
to `posterior`, one of Stan's R packages.
Along the way, we will show how you can use the Stan skills to
support your contribution workflow.

Note that these skills are still a prototype.
Please use them with care and report any problems you run into, so that we can
learn how to improve them.

You can get the skills from [https://github.com/florence-bockting/stancon-contributing](https://github.com/florence-bockting/stancon-contributing).
Fork and clone the repository.
Open the repository of interest (in this case `posterior`) and save the 
stancon-contributing/skills folder under 

  + `~/.claude/skills` (should also be read by Cursor)
  + `~/.codex/skills`
  + `~/.copilot/skills`

depending on your LLM. Alternatively, use your LLM and prompt it: 
```
read the skills from <path>/stancon-contributing/skills into your global skill
```
Even if you don't use an LLM, the skills are useful on their own: they provide
recipes for the tasks that come up again and again when contributing to an R package.

## Writing an issue {#sec-using-skills}

Suppose you hit a problem in the `posterior` package: all columns
of the `summarise_draws()` output are converted to character, even the
`mean` column, which should be numeric.

```{r, eval = FALSE}
example_draws() |>
   summarise_draws(char_fun = \(x) "word", mean)

# A tibble: 10 × 3
#   variable char_fun mean            
#   <chr>    <chr>    <chr>           
# 1 mu       word     4.1799990610081 
# 2 tau      word     4.16356885610418
# ...
```

First, we want to report this problem to the developers of the
`posterior` package, so we write an issue.
In [Writing good Issues](#sec-issues-feature-request), we discussed what the
structure of a *good* issue looks like.

The `stan-issue-and-pr` skill helps you write an issue for a given
problem that follows this structure.
The issue template is provided in [`skills/stan-issue-and-pr/templates/bug-report.md`](https://github.com/florence-bockting/stancon-contributing/blob/main/skills/stan-issue-and-pr/templates/bug-report.md).

Next, we show how the skill can be used to create an issue.

### Duplicate issue

Point the AI assistant at the skill and make sure it actually reads it.

::: {.chat}
::: {.turn .you}
[You]{.speaker}
```
I want to create an issue in posterior for the following problematic case:

example_draws() |>
   summarise_draws(char_fun = \(x) "word", mean)

Problem: all columns are converted to string even those that should be a double (e.g., mean).
```
:::
:::

The first thing the LLM does is check for duplicates.
You can look up all steps the skill asks for in [`skills/stan-issue-and-pr/SKILL.md`](https://github.com/florence-bockting/stancon-contributing/blob/main/skills/stan-issue-and-pr/SKILL.md).

In our case, an issue for this problem already exists, so the skill flags it as a
duplicate.

::: {.chat}
::: {.turn .ai}
[Claude Opus]{.speaker}
I read `skills/stan-issue-and-pr/SKILL.md` and its bug-report template. Step one
there is a duplicate search. Your case is already reported.

[stan-dev/posterior#355](https://github.com/stan-dev/posterior/issues/355) — "In
summarise draws, a function that returns a character changes all columns to
character". Open since 2024-03-14, labelled `bug`, 13 comments. It uses your
exact example:

```r
example_draws() |>
   summarise_draws(char_fun = \(x) "word", mean)
```

It also proposes a fix: `type.convert(out, as.is = TRUE)` on the output tibble
in `summarise_draws_helper`.

So a new issue would be a duplicate. Two options instead:

- I read the 13 comments and tell you where the discussion landed.
- You add a comment to #355 if you have something new — a different failure
  mode, or a reason the proposed fix does not work.

Which would you like? Also: were you planning to fix this, or only to report it?
:::
:::

### Drafting an issue

To see the rest of the workflow, we tell the LLM to skip the first step
(i.e., the search for pre-existing issues).

::: {.chat}
::: {.turn .you}
[You]{.speaker}
```
I want to create an issue in posterior for the following problematic case:

example_draws() |>
   summarise_draws(char_fun = \(x) "word", mean)

Problem: all columns are converted to string even those that should be a double (e.g., mean).

This is a demo. Don't check for existing Issues.
```
:::
:::

The skills are written so that the AI assistant acts as a mentor: it guides you
through the process step by step instead of working autonomously.
This way, you learn the contribution process yourself and keep authority over
every decision that is made.

::: {.chat}
::: {.turn .ai}
[Claude]{.speaker}
Read `stan-issue-and-pr`. This repo ships no issue template, so we use the
skill's bug-report template. We are at the submit stage — this becomes public in
your name, so [you write the words and post it]{.hl}.

I will ask the template questions one at a time and write down your answers.

**First: what did you expect?** You passed `char_fun = \(x) "word"` and `mean`.
What should the returned `draws_summary` have looked like — which columns, which
types?
:::

::: {.turn .you}
[You]{.speaker}
The mean should have been a numeric.
:::

::: {.turn .ai .fragment}
[Claude]{.speaker}
**Next: actual output.** The issue needs the real printout as text, not a
description of it. Can you paste what R printed for that call?

If you'd rather, I can run it here and show you the output — you check it, then
it goes in the issue.
:::

::: {.turn .you}
[You]{.speaker}
Run it yourself.
:::

::: {.turn .ai}
[Claude]{.speaker}
Confirmed. mean comes back as <chr>, values stringified:

```
# A tibble: 10 × 3
   variable char_fun mean            
   <chr>    <chr>    <chr>           
 1 mu       word     4.1799990610081 
 2 tau      word     4.16356885610418
...
   variable    char_fun        mean 
"character" "character" "character"
```

**Environment:** which `posterior` version did you hit this on — installed
1.7.0, or your local checkout?
:::

::: {.turn .you}
[You]{.speaker}
The version from my local checkout.
:::

::: {.turn .ai .fragment}
[Claude]{.speaker}
Two sections left: **Additional context** and **Open questions**.

Do you want me to trace where the coercion happens in `R/` and add it as a
pointer?
:::

::: {.turn .you}
[You]{.speaker}
Yes, trace it back.
No further open questions.
:::

::: {.turn .ai}
[Claude]{.speaker}
Draft written to `issue-draft.md`:
:::
:::

::: {.artifact}
[▤ issue-draft.md]{.filename}
````markdown
## What happened

`summarise_draws()` returned every summary column as character when one of the
supplied functions returned a string. `mean` came back as `<chr>` with
stringified values.

## What I expected

`mean` stays numeric (double). Only the column produced by the
character-returning function should be character.

## Minimal reproducible example

```r
library(posterior)

example_draws() |>
  summarise_draws(char_fun = \(x) "word", mean)
```

## Actual output

```
# A tibble: 10 × 3
   variable char_fun mean
   <chr>    <chr>    <chr>
 1 mu       word     4.1799990610081
 2 tau      word     4.16356885610418
 3 theta[1] word     6.74893947963962
 4 theta[2] word     5.25331634990726
 5 theta[3] word     3.04393475572924
 6 theta[4] word     4.85842854299997
 7 theta[5] word     3.22258991784476
 8 theta[6] word     3.98696993634695
 9 theta[7] word     6.5030995214442
10 theta[8] word     4.56520199862817

> sapply(x, class)
   variable    char_fun        mean
"character" "character" "character"
```

## Steps to reproduce

1. `library(posterior)`
2. Run `example_draws() |> summarise_draws(char_fun = \(x) "word", mean)`
3. Check the column types, e.g. with `sapply(x, class)`

## Environment

- OS: Ubuntu 24.04.4 LTS
- Package version: posterior 1.7.0 (local checkout at commit 4ec8b00)
- R version: 4.6.1 (2026-06-24)

## Additional context

The summary is accumulated in a single matrix rather than per column, in
`summarise_draws_helper()`:

- `R/summarise_draws.R:360` — `out <- matrix(NA, ...)` holds the columns of
  every summary function
- `R/summarise_draws.R:362` and `R/summarise_draws.R:367` — `unlist()` on a list
  mixing a character with doubles yields a character vector, and assigning it
  coerces the whole matrix
- `R/summarise_draws.R:370` — `as_tibble()` then sees character throughout

A matrix holds one type, so a single character-returning function converts all
columns.

An AI assistant helped me reproduce this and locate the code path. I checked the
output and the code references myself.
````
:::

Now suppose we want to contribute a fix for the issue above.
The instructions for the contribution workflow are described in
[`skills/stan-r-package-contribution/SKILL.md`](https://github.com/florence-bockting/stancon-contributing/blob/main/skills/stan-r-package-contribution/SKILL.md).

You can either work through the contribution yourself, using the skill as a
guideline (or checklist), or you can work through it as a dialogue with an LLM.

## Is the issue unclaimed?

::: {.chat}
::: {.turn .you}
[You]{.speaker}
I want to contribute a solution to the issue: https://github.com/stan-dev/posterior/issues/355

How do I get started?
:::

::: {.turn .ai}
[Claude]{.speaker}
I read the skills in skills/ — stan-contributing and stan-r-package-contribution govern what follows.

**Stage: orient — step 1, confirm the issue is unclaimed.**

Your checklist:
```
Orient    1. Issue read, unclaimed, understood
Set up    2. Fork and clone   3. Preflight   4. Branch
Change    5. Write code       6. Document
Verify    7. Test  8. NEWS.md  9. devtools::check()
Submit   10. Commit, push, pull request
Stage: orient — step 1, confirm the issue is unclaimed.
```
Two things from the read-only lookups you should see:

#355 already has an open PR, #356 by n-kall, last updated 2025-03-03. Claimed, but 17 months quiet.
The thread rejected the issue author's own proposal. paul-buerkner and mjskay agreed on binding results into a data frame directly, instead of a matrix or type.convert(). Also unlist() flattens types — purrr::flatten was floated, with the requirement that posterior not depend on purrr.
So the work here is not "apply type.convert()". Building that is the common way a good PR gets sent back.

Before any code, ask in the issue whether the PR is stale. Draft:
```
Hi! Is this still being worked on? #356 hasn't moved since March 2025. 
Happy to pick it up — should I build on that branch or start fresh?
```
Post it yourself, or shall I?
:::

::: {.turn .you}
Skip this step and start with fresh local PR
:::
:::

## Check setup

First, we ensure that we have a fork and clone of the repo of interest.

```
# from R
usethis::create_from_github("stan-dev/posterior", fork = TRUE)

# from the shell, clones into ./posterior
gh repo fork stan-dev/posterior --clone
```

Then, we check whether the environment is set up as expected. 
For this, the skills ship an R script,
`stan-r-package-contribution/scripts/check_environment_setup.R`, that runs a few
automatic checks.

::: {.chat}
::: {.turn .ai}
[Claude]{.speaker}
**Stage: set up — step 3, check the environment setup.**

This script reads DESCRIPTION, your R library, and your git state. It answers one question: can this machine make a clean pull request? The main check is the roxygen2 pin — a mismatch turns your 4-file change into a 47-file diff at step 6. It changes nothing. It prints one OK/WARN/FAIL line per check.

```
Rscript skills/stan-r-package-contribution/scripts/check_environment_setup.R
```

Run it, or shall I? Paste what it prints.
:::

::: {.turn .you}
[You]{.speaker}

```bash
Rscript skills/stan-r-package-contribution/scripts/check_environment_setup.R
```

```
Environment check for an R package contribution.
Reads DESCRIPTION, your R library, and your git state. Changes nothing.
FAIL blocks a reviewable pull request. WARN is worth a look.

[OK  ] Package: posterior
[OK  ] Contributing guide: .github/CONTRIBUTING.md -- read it before you start.
[FAIL] roxygen2 mismatch: installed 8.0.0, DESCRIPTION pins 7.3.3.
       Documenting now would rewrite every file in man/.
       remotes::install_version("roxygen2", version = "7.3.3")
[OK  ] devtools installed.
[OK  ] usethis installed.
[OK  ] testthat installed.
[OK  ] vdiffr installed (this package uses snapshot tests).
[OK  ] Working tree clean.
       Branch 'add-ai-contribution-policy' is not a base branch.
       Did you create it for this contribution? [y/N] N

[FAIL] Branch 'add-ai-contribution-policy' is not for this contribution.
       Your change would start from somebody else's work.
       git checkout main   # or master, or develop
       usethis::pr_init("123-short-description")
[OK  ] gh CLI authenticated.

Fix every FAIL above before you start the change.
```
:::
:::

::: {.callout-note}
## If `Rscript` is not found

On Windows the R installer does not add `R\bin` to your `PATH`, so this command
fails with "not recognized" before it checks anything. Either add `R\bin` to the
`PATH`, or skip the script and run the same checks by hand from the R console —
`packageVersion("roxygen2")` against the pin in `DESCRIPTION` is the one that
matters. The script is a convenience; the checks are the point.

One check further ahead has a platform requirement too: `devtools::check()`
compiles the package, which needs
[RTools](https://cran.r-project.org/bin/windows/Rtools/) on Windows or the Xcode
command line tools (`xcode-select --install`) on macOS. Sort that out here, at
the set-up stage, rather than after the change is written.
:::

## Code contributions

The next step is to make the code changes.
This step should always consist of two connected parts: (1) a test that checks
the outcome you want to achieve, and (2) the code changes needed to achieve it.

In our example, we want `summarise_draws()` to return an output in which the type
of one column is not carried over to the other columns.

So we could, for example, write the following tests into
`tests/testthat/test-summarise_draws.R`.

```{r, eval=FALSE}
test_that("summarise_draws_helper works correctly", {
  res1 <- example_draws() %>%
    summarise_draws(char_fun = function(x) "word", mean)
  
  expect_type(res1$char_fun, "character")
  expect_type(res1$mean, "double")

  res2 <- example_draws() %>%
    summarise_draws(char_fun = function(x) factor("a"), mean)
  
  expect_s3_class(res2$char_fun, "factor")
  expect_type(res2$mean, "double")

  res3 <- example_draws() %>%
    summarise_draws(char_fun = function(x) TRUE, mean)
  
  expect_type(res3$char_fun, "logical")
  expect_type(res3$mean, "double")
})

```

Then we make the code changes needed to reach the desired output.
In this case, we change a helper function that is used by
`summarise_draws()` in `R/summarise_draws.R`:

```{r, eval=FALSE}
# flatten the summary list of a single variable into a list with one element
# per output column, keeping each value in its original type
flatten_summary_list <- function(v_summary) {
  unlist(lapply(v_summary, as.list), recursive = FALSE, use.names = FALSE)
}

summarise_draws_helper <- function(x, funs, .args) {
  variables_x <- variables(x)
  # get length and output names, calculated on the first variable
  out_1 <- create_summary_list(x, variables_x[1], funs, .args)
  the_names <- vector(mode = "list", length = length(funs))
  for (i in seq_along(out_1)){
    if (rlang::is_named(out_1[[i]])) {
      the_names[[i]] <- names(out_1[[i]])
    } else if (length(out_1[[i]]) > 1) {
      the_names[[i]] <- paste0(names(out_1)[i], ".", c(1:length(out_1[[i]])))
    } else {
      the_names[[i]] <- names(out_1)[i]
    }
  }
  the_names <- unlist(the_names)
  # Check for naming issues prior do doing lengthy computation
  if ("variable" %in% the_names) {
    stop_no_call("Name 'variable' is reserved in 'summarise_draws'.")
  }
  # Pre-allocate list to store output, one element per variable holding
  # the summary values of that variable as a flat (but untyped) list
  out <- vector("list", length(variables_x))
  out[[1]] <- flatten_summary_list(out_1)
  # Do the computation for all remaining variables
  if (length(variables_x) > 1L) {
    for (v_ind in 2:length(variables_x)) {
      out_v <- create_summary_list(x, variables_x[v_ind], funs, .args)
      out[[v_ind]] <- flatten_summary_list(out_v)
    }
  }
  # bind the summaries column-wise so that the type of each summary is
  # preserved instead of being coerced to a common type across summaries
  columns <- lapply(seq_along(the_names), function(i) {
    vctrs::vec_c(!!!lapply(out, .subset2, i))
  })
  names(columns) <- the_names
  out <- tibble::as_tibble(columns)
  out$variable <- variables_x
  out <- move_to_start(out, "variable")
  class(out) <- class_draws_summary()
  out
}
```

Finally, we run the tests to check whether our code changes make them pass.

```bash
testthat::test_file("tests/testthat/test-summarise_draws.R")
```

## Wrapping up

Once the code changes and tests are done, a few final tasks remain before we can
open a pull request. For orientation, we can again check
`skills/stan-r-package-contribution/SKILL.md` or simply ask the LLM:

::: {.chat}
::: {.turn .you}
[You]{.speaker}
Changes are made. Is this branch PR ready?
:::

::: {.turn .ai}
[Claude]{.speaker}
**Stage: verify — the branch is not PR ready yet.**

What is missing:

Nothing is committed. git log master..HEAD is empty. Both files are still working-tree changes only.
No NEWS.md entry (step 8). The file starts at # posterior 1.7.1, so you need a new # posterior (development version) header with a ### Bug Fixes bullet citing (#355).
devtools::check() has not been run (step 9).
No fork remote. origin is stan-dev/posterior itself. If you have push rights that is fine; otherwise usethis::pr_push() will fail.

Next step is NEWS.md. The skill is firm that you write this one — it is the only text users read under the package name.

Shall I show you the recent entries to match the style, or do you want to draft it straight away?
:::
:::

### Update NEWS.md

The `NEWS.md` file keeps track of all changes made to the package over time.
When a new release is published, it is the reference used to communicate which
changes that release contains.

::: {.artifact}
[▤ NEWS.md]{.filename}

```
# posterior (development version)

* Ensure `summarise_draws()` preserves the type of each summary, so that a
function returning a character or factor no longer coerces the numeric
columns to character (#355).

...

```
:::

### Run CRAN checks

Run the same checks that are run when a package is released to CRAN. This is a
larger test suite that builds the vignettes and checks the documentation, the
examples, and more. Depending on the package, this can take a while.

::: {.chat}
::: {.turn .ai}
[Claude]{.speaker}

Then step 9. devtools::check() runs what CRAN and the CI run: it rebuilds the docs, runs the full suite in a clean session, and checks the examples and NEWS.md formatting. Finding a NOTE here costs you minutes; finding it in review costs you days. It takes several minutes.

```bash
devtools::check()
```

Run it, or shall I? Paste the summary lines it ends with.

:::

::: {.turn .you}
[You]{.speaker}

*Run `check()` command.* 

```
── R CMD check ─────────────────────────────────────────────────────────────────────────────────────────────
─  using log directory ‘/tmp/RtmpcegVNv/filefd6f13521fdce/posterior.Rcheck’
─  using R version 4.6.1 (2026-06-24)
─  using platform: x86_64-pc-linux-gnu
─  R was compiled by
       gcc (Ubuntu 13.3.0-6ubuntu2~24.04.1) 13.3.0
       GNU Fortran (Ubuntu 13.3.0-6ubuntu2~24.04.1) 13.3.0
─  running under: Ubuntu 24.04.4 LTS
─  using session charset: UTF-8
─  current time: 2026-08-16 04:09:51 UTC
─  using options ‘--no-manual --as-cran’
✔  checking for file ‘posterior/DESCRIPTION’ ...
─  this is package ‘posterior’ version ‘1.7.1’
─  package encoding: UTF-8
✔  checking package namespace information ...
─  checking package dependencies ... INFO (5.1s)
   Packages suggested but not available for checking:
     'caret', 'gbm', 'randomForest', 'e1071'
✔  checking if this is a source package ...
✔  checking if there is a namespace
✔  checking for executable files (381ms)
✔  checking for hidden files and directories ...
✔  checking for portable file names ...
✔  checking for sufficient/correct file permissions ...
✔  checking serialization versions ...
✔  checking whether package ‘posterior’ can be installed (3.4s)
✔  checking installed package size ...
✔  checking package directory
✔  checking for future file timestamps (395ms)
✔  checking ‘build’ directory ...
✔  checking DESCRIPTION meta-information (406ms)
N  checking top-level files
   Non-standard file/directory found at top level:
     ‘skills’
✔  checking for left-over files
✔  checking index information ...
✔  checking package subdirectories (730ms)
✔  checking code files for non-ASCII characters ...
✔  checking R files for syntax errors ...
✔  checking whether the package can be loaded ...
✔  checking whether the package can be loaded with stated dependencies ...
✔  checking whether the package can be unloaded cleanly ...
✔  checking whether the namespace can be loaded with stated dependencies (338ms)
✔  checking whether the namespace can be unloaded cleanly (373ms)
✔  checking loading without being on the library search path (365ms)
✔  checking whether startup messages can be suppressed (692ms)
✔  checking dependencies in R code (495ms)
✔  checking S3 generic/method consistency (530ms)
✔  checking replacement functions ...
✔  checking foreign function calls (494ms)
✔  checking R code for possible problems (7.1s)
✔  checking Rd files (592ms)
✔  checking Rd metadata ...
✔  checking Rd line widths ...
✔  checking Rd cross-references (450ms)
✔  checking for missing documentation entries (382ms)
✔  checking for code/documentation mismatches (1.2s)
✔  checking Rd \usage sections (876ms)
✔  checking Rd contents ...
✔  checking for unstated dependencies in examples ...
✔  checking R/sysdata.rda ...
✔  checking installed files from ‘inst/doc’ ...
✔  checking files in ‘vignettes’ ...
✔  checking examples (7.1s)
✔  checking examples with --run-donttest (6.4s)
✔  checking for unstated dependencies in ‘tests’ ...
─  checking tests ...
✔  Running ‘testthat.R’ (8s)
✔  checking for unstated dependencies in vignettes (452ms)
✔  checking package vignettes (934ms)
✔  checking re-building of vignette outputs (5.7s)
✔  checking for non-standard things in the check directory ...
✔  checking for detritus in the temp directory
   
   See
     ‘/tmp/RtmpcegVNv/filefd6f13521fdce/posterior.Rcheck/00check.log’
   for details.
   
── R CMD check results ──────────────────────────────────────────────────────────────── posterior 1.7.1 ────
Duration: 57.4s

❯ checking top-level files ... NOTE
  Non-standard file/directory found at top level:
    ‘skills’

0 errors ✔ | 0 warnings ✔ | 1 note ✖
```
:::
:::

Everything is fine. The only complaint is that the `skills` folder is flagged as
a non-standard directory. Adding the line `^skills$` to `.Rbuildignore` makes
this *note* disappear.

#### Add, commit, and push your changes (open your PR)
Next, add, commit, and push the changes with git.
Commit in reasonable chunks and make sure that no unintended changes end up in
the pull request. This is not always easy, especially for larger refactorings.
Commit histories can be cleaned up afterwards (e.g., by squashing commits), but
for the Stan R packages we would not recommend it: it is extra effort for the
contributor without much benefit. Bottom line: try to work methodically and
commit cleanly, but don't panic if your history is a bit messy. Just leave it as is.

```bash
git add R/summarise_draws.R
git commit -m "fix: summarise_draws_helper collects summary values column-wise"

git add tests/testthat/test-summarise_draws.R
git commit -m "tests: add tests for updated summarise_draws_helper"

git add NEWS.md
git commit -m "chore: update NEWS.md"
```

Finally, open a pull request with

```bash
usethis::pr_push()
```

## PR description
The final task is to write the pull request description.
The [`pr-description.md`](https://github.com/florence-bockting/stancon-contributing/blob/main/skills/stan-issue-and-pr/templates/pr-description.md) template helps you with this.

Following that template, a description for our example could look as follows:

::: {.artifact}
[▤ pr-description.md]{.filename}

````markdown
Closes #355

## Description

`summarise_draws()` collected all summary values in a matrix, which holds one
type, so a character-valued summary function coerced every column to character.
The values are now collected in a list per variable and bound column-wise with
`vctrs::vec_c()`, so each column keeps the type its function returned. This
follows the approach agreed in the issue thread.

## Breaking changes

A summary function whose return type differs across variables now raises an
error from `vec_c()` instead of coercing to character.

## Tests

Added a test that summary functions returning character, factor, and logical
leave the `mean` column `double`.

    [ FAIL 0 | WARN 0 | SKIP 11 | PASS 2032 ]
    devtools::check(): 0 errors | 0 warnings | 0 notes

## Documentation

`NEWS.md` bullet added. No user-facing docs changed; the new helper is internal.

## Open questions / Uncertainties

- I added the `skills/` folder to `.Rbuildignore`. If this is not desired, I can revert that change.

## AI assistance

Claude helped with drafting this PR. I reviewed every change and can
explain it.

## Copyright and Licensing

Copyright holder: USER or UNIVERSITY

By submitting this pull request, the copyright holder agrees to license the
submitted work under the following licenses:

- Code: BSD 3-clause (https://opensource.org/licenses/BSD-3-Clause)
- Documentation: CC-BY 4.0 (https://creativecommons.org/licenses/by/4.0/)

## Checklist

- [x] Style is consistent with the existing code and docs
- ~[ ] Documentation or vignette renders locally~
- [x] Tests pass, with the output pasted above
- [x] Full check passes (`devtools::check()` / `pytest` + `ruff`)
- [x] Changelog handled — `NEWS.md` entry added, or the title is
      changelog-worthy where the changelog is generated
- [x] Copyright holder named
- [x] Every file in the diff is explainable
- [x] AI assistance disclosed, per the
      [AI Contribution Policy](https://github.com/stan-dev/stan/wiki/AI-Contribution-Policy)
````
:::
